---
title: "Reproducible Research - Project 1"
author: "Vinicius Pereira"
date: "November 10th, 2017"
output: md_document
variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading data

```{r, warning=FALSE,message=FALSE}
library(dplyr)
setwd("c:/r_workspace/rep_research")
# read dataset from activity.csv file.
activityDataset <- read.csv('activity.csv')
# converts the variable date from string to date.
activityDataset$date <- as.Date(activityDataset$date, "%Y-%m-%d")

# Summarize the data into a new dataframe (mean,sum and median) grouped by date variable
summarizedData <- activityDataset %>% group_by(date) %>%
          summarize(meanSteps = mean(steps, na.rm = T)
                    ,totalSteps = sum(steps, na.rm = T)
                    ,medianSteps = median(steps, na.rm = T))

head(summarizedData)
```

## Histogram of the total number of steps taken each day 
```{r}
# Plot summarized data histograms
par(mfrow=c(1,3))
hist(summarizedData$totalSteps, main="Total steps each day",xlab = "Total steps per day",ylab = "Frequency")
hist(summarizedData$meanSteps, main="Mean steps per day",xlab = "Mean steps per day",ylab = "Frequency")
hist(summarizedData$medianSteps, main="Median steps per day",xlab = "Median steps per day",ylab = "Frequency")
```

## Time series plot of the average number of steps taken
```{r}
plot(summarizedData$date,summarizedData$meanSteps,xlab="Date",ylab="Mean steps",type="l")
```

## The 5-minute interval that, on average, contains the maximum number of steps
```{r}
# Get the mean of steps grouping by the variable *interval*
stepsAverageMax <- activityDataset %>% group_by(interval) %>%
     summarize(meanSteps = mean(steps, na.rm = T))
# print a sample of the dataset containg the interval and the mean steps
head(stepsAverageMax)
plot(stepsAverageMax$interval,stepsAverageMax$meanSteps,xlab="interval",ylab="Mean steps",type="l")
```

The period with higer number of steps in the observation is between interval 750 and interval 1000.
The interval 835 is peak, containing the maximum number of steps.
```{r}
maxInterval <-  stepsAverageMax[order(-stepsAverageMax$meanSteps),]
head(maxInterval,5)
```

## Imputing missing data
In order to clean the dataset and reduce the bias into calculation and summarized data, NAs will be substituted by 0. 
```{r}
activityDatasetNARemoved <- activityDataset
#Substitute all NAs values with 0
activityDatasetNARemoved[is.na(activityDatasetNARemoved)] <- 0

#histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
totalStepsNARemoved <- activityDatasetNARemoved %>% group_by(interval) %>%
     summarize(totalSteps = sum(steps, na.rm = T))

# Plot the comparison between raw activity data ans data with NAs removed
par(mfrow=c(1,2))
hist(summarizedData$totalSteps, main="Total steps each day",xlab = "Total steps per day",ylab = "Frequency")
hist(totalStepsNARemoved$totalSteps, main="Total steps each day\n(NAs removed)",xlab = "Total steps per day",ylab = "Frequency")

# Create a dataset with Mean and Median for raw activity data and NAs removed
comparisonDataset <- c(mean(activityDataset$steps),median(activityDataset$steps),mean(activityDatasetNARemoved$steps),median(activityDatasetNARemoved$steps))

comparisonDataset
```
The first two values *NA* are reffered with the raw activity data, with NA's. The last two values are the mean and median considering the NAs removed from the dataset.


## Average number of steps taken per 5-minute interval across weekdays and weekends 
```{r echo=FALSE,results=FALSE}
Sys.setlocale("LC_TIME", "C")
```
```{r}
# Get day of week name from *date* in dataset and set if it is weekend or weekday
activityDatasetNARemoved$day <- ifelse(weekdays(activityDatasetNARemoved$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

#Subset data based on day variable (weekday or weekend)
weekdaysActivity <- activityDatasetNARemoved[activityDatasetNARemoved$day == "weekday",]
weekendDaysActivity <- activityDatasetNARemoved[activityDatasetNARemoved$day == "weekend",]

#Summarize the data (get mean) grouping by interval variable
weekdaysActivityMeanSteps <- weekdaysActivity %>% group_by(interval) %>%
          summarize(meanSteps = mean(steps, na.rm = T))

weekendActivityMeanSteps <- weekendDaysActivity %>% group_by(interval) %>%
          summarize(meanSteps = mean(steps, na.rm = T))

#print the comparison between weekend and weekdays activity
plot(weekendActivityMeanSteps$interval,weekendActivityMeanSteps$meanSteps,xlab="Interval",ylab="Mean steps",type="l", main="Average steps per interval\nAccros weekdays and weekend",col="red",ylim=range(0:200))
lines(weekdaysActivityMeanSteps$interval,weekdaysActivityMeanSteps$meanSteps, type="l", col="blue")
legend("topright",c("weekend","weekday"),col=c("red","blue"),pch=21)
axis(1,c(0,250,500,750,1000,1250,1500,1750,2000,2250,2500))
axis(2,c(0,25,50,75,100,125,150,175,200))
```

In the line chart above, it is clearly that in weekend days the activity starts later than in weekend days and intensifies during period from 1000 and 2000. In the other hand, activity in weed days starts around 500 in average and has the peak in the 835 interval, ranges over the period from 1250 to 2000.